---
title: "Introduction to microSTASIS"
author: 
  - name: Pedro Sánchez Sánchez
    email: bio.pedro.technology@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('microSTASIS')`"
vignette: >
  %\VignetteIndexEntry{Introduction to microSTASIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()
## Bib setup
library("RefManageR")
## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    microSTASIS = citation("microSTASIS")[1]
)
```

# Introduction

The initial idea behind **µSTASIS** was to warrant the proper processing of high-dimensional, sparse, undetermined, right skewed, overdispersed and compositional data. Specifically, it aims to fill the gap of temporal stability metrics in the compositional analysis of microbiome data. However, it can also work if the data are not transformed for belonging to the [Aitchison simplex](https://en.wikipedia.org/wiki/Compositional_data).

On one side, the assumptions in which µSTASIS is based are two: the individual-specific composition and the within-individual variability over time. On the other side, the output, which we called mS score, is easy to interpret and provides a contextualized and intuitive metric to estimate temporal microbiota stability. Also, the package incorporates cross-validation routines (leave-one-out and k-fold), that compute the mean absolute error,  and multiple functions to visualize the result.

Therefore, sort of the questions that could be answered are related with robust time-resolved definitions of stability to assess microbiota behaviour against perturbations or to search for specific compositions that present a dynamic equilibrium around a central attractor state rather than overcoming a threshold of no return.

## Algorithm

Firstly, two samples from the same individual has to be paired. For that, one can merge the sequential paired times (t1 with t2, t2 with t3...) or use a non-sequential order (t1 with t3). Therefore, from a single large data matrix, we would generate more that are smaller, that is, with fewer observations (samples).

Once we have the paired samples, it is time for the main algorithm: iterative clustering. Concretely, Hartigan-Wong k-means algorithm is used as many times as possible for stressing out paired samples from the same individuals to test if they remain together for multiple numbers of clusters over a whole data set of individuals. Also, this is corrected in those cases where the distance between samples from the same individuals is lower than the distances to their respective centroids.

```{r, echo = FALSE}
image <- "https://github.com/BiotechPedro/microSTASIS/blob/main/man/figures/microSTASIS_algo.png"
cat(paste0('<center><img src="', image,  '"></center>')) 
```

* `r Biocpkg("microSTASIS")` `r Citep(bib[["microSTASIS"]])`

# Basics

## Install `microSTASIS`

[comment]: <> (`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages.) 
`r Biocpkg("microSTASIS")` is a `R` package available via the [Bioconductor](http://bioconductor.org) repository (formerly at [CRAN](https://cran.r-project.org/package=microSTASIS)) which can be installed by using the following commands in your `R` session:

[comment]: <> (`R` can be installed on any operating system from [CRAN](https://cran.r-project.org/) after which you can install `r Biocpkg("MyBioconductorPackage")`)

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
  }
BiocManager::install("microSTASIS")
## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

[comment]: <> (
## Required knowledge

`r Biocpkg("microSTASIS")` is based on many other packages and in particular in those that have implemented the infrastructure needed for dealing with microbiome data, such as `r Biocpkg("SummarizedExperiment")` or `r Biocpkg("TreeSummarizedExperiment")`.

If you are asking yourself the question "Where do I start using Bioconductor?" you might be interested in [this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `microSTASIS` tag and check [the older posts](https://support.bioconductor.org/t/microSTASIS/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.


## 
)

## Citing `microSTASIS`

We hope that `r Biocpkg("microSTASIS")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("microSTASIS")
```

# Quick start to using `microSTASIS`

```{r "start", message=FALSE}
library("microSTASIS")
```
The first step is to subset an initial matrix of microbiome data with multiple sampling points. The`pairedTimes()` function already do it for every possible paired times in a `sequential = TRUE` way or for specific times points, for example: `sequential = FALSE, specifiedTimePoints = c("1", "3")`. The output is a list with length equal to the number of paired times.

```{r "pairedTimes"}
times <- pairedTimes(data = clr, sequential = TRUE, common = "_0_")
```

Then, it is time for `iterativeClustering()`, which can be done in parallel and runs for every item in the list.

```{r "iterativeClustering"}
mS <- iterativeClustering(pairedTimes = times, parallel = TRUE, common = "_0_")
```

## Visualization

There are two main functions for visualizing the results: a scatter plot with boxplots on the side and a heatmap.

```{r "visualization"}
results <- mSpreviz(results = mS, times = times)
mSscatter(results = results, order = "median", times = c("t1_t25", "t25_t26"), gridLines = TRUE, sideScale = 0.2)
mSheatmap(results = results, order = "mean", times = c("t1_t25", "t25_t26"), label = TRUE)
```

## Cross-validation

The stability results can be validated by cross validation (CV) in two ways: leave-one-out or k-fold. `k` is the number of individuals removed in each time that `iterativeClustering()` is run internally; it is equal to 1 for the LOOCV.

```{r "CV"}
cv_klist_k2 <- BiocParallel::bpmapply(iterativeClusteringCV, name = names(times), k = rep(2L, 3), 
                                      MoreArgs = list(pairedTimes = times, results = mS, common = "_0_"), 
                                      BPPARAM = BiocParallel::bpparam())
```

Then, the result can be displayed in two ways: as the mean absolute error or as a visual representation of how the mS score change.

```{r "error"}
MAE_t1_t25 <- mSerrorCV(pairedTime = times$t1_t25, CVklist = cv_klist_k2[[1]],  k = 2L)
MAE <- mSpreviz(results = list(MAE_t1_t25), times = list(t1_t25 = times$t1_t25))
mSheatmap(results = MAE, times = c("t1_t25", "t25_t26"), label = TRUE,
          high = 'red2',  low = 'forestgreen', midpoint = 5)
```

```{r "lines"}
mSlinesCV(pairedTime = times$t1_t25, CVklist = cv_klist_k2[[1]], k = 2L)
```

## Metadata integration

The analysis can also be integrated with metadata of the individuals in the following way.

```{r "metadata"}
metadata <- data.frame(Sample = rownames(clr), age = c(rep("youth", 65), rep("old", 131-65)))
group <- mSmetadataGroups(metadata = metadata, samples = metadata$Sample, common = "_0_", 
                          individuals = results$individual, variable = "age")
mSdynamics(results, groups = group, points = TRUE, linetype = 0)
```

## Others

This example run on using a matrix as input but the package allows to easily use a TreeSummarizedExperiment object, as can be seen in the example code for each `help()` section of the functions.

# Reproducibility

The `r Biocpkg("microSTASIS")` package `r Citep(bib[["microSTASIS"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("microSTASIS.Rmd", "BiocStyle::html_document"))
## Extract the R code
library("knitr")
knit("microSTASIS.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```